A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-12-05 Simulation/Writeup/NoteConvergence.Rnw
Added new graphs to manuscript (badly)
some spelling errors

At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}. Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., listwise deletion), which can yield wildly invalid results \citep{buur18}. A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}. MI is an iterative algorithmic procedure in which each missing data point is `guessed' (i.e. imputed) several times. The variablity between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness. Therefore, MI can provide valid inferences despite of missing information.

\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be correct. If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it requires more observations than strictly required. Since both of these situations are undesirable, imputations and their variability should be properly evaluated. Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice}. 
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
The aim of this note is to provide guidelines for the evaulation of imputations to applied researchers who use MI methods. More specifically, this note adresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.

@@

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability to be missing is equal for all $n \times p$ cells in $Y$.

In this note, the terms `unobserved' and `missing' data are used interchangibly. Both refer to $Y_{mis}$, e.g. `$NA$' values in \proglang{R} or `999' in \proglang{SPSS}. The terms 'incomplete' or 'observed' data denote $Y_{obs}$. Incomplete data is the starting point of the multiple imputation procedure. Figure \ref{fig:steps} provides a schematic overview of the steps involved with MI.

@@

\textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
Explain that there is no baseline to compare the measures with. Only current convergence check is to visually inspect trace plots. (But in the traceplots convergence seems immediate).}

@@

Autocorrelation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration. For now, we only consider the chain mean as scalar of interest $\theta$. The AC of the variable with the highest absolute AC value is reported per simulation. These values are then averaged per simulation condition.

@@

Confidence interval width (CIW) is computed as the diffence between the lower and upper bound of the 95\% confidence interval (CI95\%). The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom). CIW is averaged over all simulations with the same simulation condition.

@@

\textbf{Try to put the chain means and vars together in one panel}

% \begin{figure}[h]
%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
%         \includegraphics{Figures/results.pdf}
%   }
%   \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
%     \label{resultssim}
% \end{figure}

@@

\textbf{$\widehat{R}$}. Steep drop down up-to iteration twenty, then more gradual decrease upto iteration fourty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations (see online appendix for full table of results).

@@

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} From two iterations upwards, average bias across repititions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending upto fourty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.

@@

From the simulation diagnostics it appears that as little as two iterations could be sufficient to draw valid inference. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate deviant behavior upto twenty or even fourty iterations.

>>>>

At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}. Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which can yield wildly invalid results \citep{buur18}. A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}. MI is an iterative algorithmic procedure in which each missing data point is `guessed' (i.e. imputed) several times. The variability between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness. Therefore, MI can provide valid inferences despite of missing information.

\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be correct. If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it requires more observations than strictly required. Since both of these situations are undesirable, imputations and their variability should be properly evaluated. Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice}. 
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
The aim of this note is to provide guidelines for the evaluation of imputations to applied researchers and statisticians who use MI methods. More specifically, this note addresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.

@@

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.

In this note, the terms `unobserved' and `missing' data are used interchangeably. Both refer to $Y_{mis}$, e.g. `$NA$' values in \proglang{R} or `999' in \proglang{SPSS}. The terms 'incomplete' or 'observed' data denote $Y_{obs}$. Incomplete data is the starting point of the multiple imputation procedure. Figure \ref{fig:steps} provides a schematic overview of the steps involved with MI.

@@

\textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
Explain that there is no baseline to compare the measures with. Only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}

@@

Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration. For now, we only consider the chain mean as scalar of interest $\theta$. The AC of the variable with the highest absolute AC value is reported per simulation. These values are then averaged per simulation condition.

@@

Confidence interval width (CIW) is computed as the difference between the lower and upper bound of the 95\% confidence interval (CI95\%). The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom). CIW is averaged over all simulations with the same simulation condition.

@@

\textbf{Try to put the chain means and vars together in one panel, like \ref{fig:rhat} but better.}

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{Figures/Rhat.pdf}
  }
  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
    \label{fig:rhat}
\end{figure}

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{Figures/AC.pdf}
  }
  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
    \label{fig:ac}
\end{figure}

@@

\textbf{$\widehat{R}$}. Steep drop down up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations (see online appendix for full table of results).

@@

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} From two iterations upwards, average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending up-to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.

@@

From the simulation diagnostics it appears that as little as two iterations could be sufficient to draw valid inference. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate deviant behavior up-to twenty or even forty iterations.

### 2019-12-05 Simulation/Writeup/NoteConvergence.tex
Added new graphs to manuscript (badly)
some spelling errors

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} From two iterations upwards, average bias across repititions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending upto fourty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.

>>>>

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} From two iterations upwards, average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending up-to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.
